---
title: "GLMM example 9"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../public/resources/ws_style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../public/resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(broom.mixed) #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(ggeffects) #for effects plots in ggplotjk
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(DHARMa)    #for assessing dispersion etc
library(glmmTMB)    #for glmmTMB
library(performance) #for diagnostic plots
library(see)         #for diagnostic plots
library(ordinal)    #for ordinal models
```

<div class="HIDDEN">

```{r readDataP, results='markdown', eval=TRUE}
hughes = read_csv('../public/data/hughes_full.csv', trim_ws=TRUE)
glimpse(hughes)
```

| Bleaching Score | 2016 Equates to  | 2017 Equates to  |
|-----------------+------------------+------------------|
|               0 | No bleaching     | No bleaching     |
|               1 | 0-10% bleaching  | 0-10% bleaching  |
|               2 | 10-30% bleaching | 10-30% bleaching |
|               3 | 30-60% bleaching | 30-60% bleaching |
|               4 | > 60% bleaching  | 60-80% bleaching |
|               5 | -                | > 80% bleaching  |

# Data processing

- Make a categorical version of Year
- Combine Scores 4 and 5 into a single Score of 4 to be consistent across years
- Make Score an ordered categorical variable
- Generate a numeric version of categorical Score (as required for the `ocat` family.  Note this must be a 1 indexed numeric.
- Relevel Sector such that it is in the order North, Central, South
- Ensure that there is a categorical version of Reef
- Generate a numeric version of the categorical Reef 


```{r processDataP, results='markdown', eval=TRUE}
hughes = hughes %>%
    mutate(fYear=factor(Year),
           Score=ifelse(Score==5,4,Score),
           oScore = factor(Score, ordered=TRUE),
           nScore = as.numeric(factor(Score, ordered=TRUE)),
           SectorThree=factor(SectorThree, levels=c('North','Central','South')),
           fReef=factor(ReefID),
           nReef=as.numeric(fReef))
# now make a version that is just 2016
hughes <- hughes %>% filter(fYear==2016) %>%
    dplyr::select(REEF=ReefID, HABITAT=Habitat, SECTOR=SectorThree, SCORE=Score)
write_csv(hughes, file='../public/data/hughes.csv')
## hughes.colors = c('#FFFFFF', rev(heat.colors(length(levels(hughes$oScore))))[-1])
```
</div>

# Scenario

# Read in the data

2016 data only

```{r readData, results='markdown', eval=TRUE}
hughes = read_csv('../public/data/hughes.csv', trim_ws=TRUE)
glimpse(hughes)
```

| REEF   | HABITAT | SECTOR | SCORE |
|--------|---------|--------|-------|
| 09-357 | C       | North  | 4     |
| 09-357 | F       | North  | 4     |
| 09-357 | U       | North  | 3     |
|        |         |        |       |

# Data preparation

<div class="HIDDEN">

- ensure score is ordinal
- declare categorical variables as factors - and relevel sector
- declare random effects as factors

```{r processData, results='markdown', eval=TRUE}
hughes = hughes %>%
    mutate(oSCORE = factor(SCORE, ordered=TRUE),
           HABITAT = factor(HABITAT),
           SECTOR=factor(SECTOR, levels=c('North','Central','South')),
           REEF=factor(REEF))
## hughes.colors = c('#FFFFFF', rev(heat.colors(length(levels(hughes$oSCORE))))[-1])
```
</div>

# Exploratory Data Analysis

Proportional (cumulative link) odds-ratio models are useful when the latent
(unmeasured) response is recorded on a ordinal (ordered categorical) scale.
When this is the case, we can calculate the probability of a that an observed
ordinal score ($y$) is less than or equal to any given level ($i$: category) given a set of
covariates ($\boldsymbol{X}$) according to the following:

$$ Pr(y\le i|\boldsymbol{X}) =
\frac{1}{1+e^{-(\theta_i - \boldsymbol{\beta}.\boldsymbol{X})}}\\ $$

where $y$ is the observed categorical response, $\boldsymbol{X}$ is a ($n \times
p$) effects model matrix, $\boldsymbol{\beta}$ are the $p$ effects parameters
and $\theta_i$ are the $K-1$ category thresholds

<div class="HIDDEN">

- count the number of samples in each bleaching category per sector per habitat
- express this number as proportions
- reverse the order of the score levels

```{r EDA2, results='markdown', eval=TRUE, fig.width=10, fig.height=10}
# Scatterplot
hughes %>%
    ggplot(aes(y = oSCORE, x = HABITAT)) +
    geom_point(position = position_jitter()) +
    facet_wrap(~SECTOR)
# Line plot - too messy to make much of this...
hughes %>%
    group_by(SECTOR, REEF, HABITAT) %>%
    summarise(SCORE = mean(SCORE)) %>%
    ungroup() %>%
    ggplot(aes(y = SCORE, x = as.numeric(HABITAT), group = REEF)) +
    geom_blank(aes(x = HABITAT)) +
    geom_line() +
    facet_grid(~SECTOR)


hughes %>%
    group_by(SECTOR, HABITAT, oSCORE) %>%
    summarise(n = n()) %>%
    ungroup() %>%
    group_by(SECTOR, HABITAT) %>%
    mutate(prop = n/sum(n)) %>%
    mutate(oSCORE = factor(oSCORE, levels = rev(levels(oSCORE)))) ->
    hughes.sum
    
## hughes.sum <- hughes %>%
##     count(SECTOR,HABITAT,oSCORE) %>%
##     group_by(SECTOR, HABITAT) %>%
##     mutate(prop=prop.table(n),
##            oSCORE=factor(oSCORE, levels=rev(levels(oSCORE))))

hughes.sum %>% head

ggplot(data=hughes.sum, aes(y=prop, x=HABITAT)) +
    geom_bar(stat='Identity', aes(fill=oSCORE), color='black') +
    facet_grid(~SECTOR) +
    ## scale_fill_manual('Bleaching score', values=rev(hughes.colors) ) +
    scale_fill_manual('Bleaching score', values=c(heat.colors(5)[-5], '#FFFFFF') ) +
    scale_y_continuous('Proportion of Reef', expand=c(0,0))+
    theme_bw() + 
    theme(panel.spacing.y=unit(10,'pt'))
```

</div>

# Fit models

<div class="HIDDEN">

```{r mod1Fit, results='markdown', eval=FALSE, cache=TRUE, results='hide'}
library(ordinal)
hughes.clmm=ordinal::clmm(oSCORE ~ HABITAT*SECTOR+(1|REEF), data=hughes)
hughes.clmm=clmm(oSCORE ~ HABITAT*SECTOR+(1|REEF), data=hughes)
hughes.clmm1=ordinal::clmm(oSCORE ~ HABITAT*SECTOR+(HABITAT|REEF), data=hughes)
AIC(hughes.clmm, hughes.clmm1)
allEffects(hughes.clmm) %>% plot
allEffects(hughes.clmm1) %>% plot
## plot_model(hughes.clmm, type='eff', terms=c('HABITAT','SECTOR'))
## plot_model(hughes.clmm, type='eff', terms=c('SECTOR','HABITAT'))
## ggpredict(hughes.clmm) %>% plot
hughes.clmm1 %>% ggpredict(c('HABITAT', 'SECTOR')) %>% plot
hughes.clmm1 %>% ggemmeans(~HABITAT|SECTOR) %>% plot

## predict(hughes.clmm)
## model=hughes.clmm
## coefs <- c(model$beta, unlist(model$ST))
#autoplot(hughes.clmm1, what='qq')
## simulateResiduals(hughes.clmm, plot=TRUE)
## hughes.clmm=ordinal::clmm(oScore ~ Habitat*fYear*SectorThree+(1|ReefID), data=hughes)
summary(hughes.clmm)
summary(hughes.clmm1)
## Habitat F is associated with a lower prob of higher bleaching - but not significant
## Habitat L is associated with a lower prob of higher bleaching (In the North)
exp(-1.655)
## Habitat U is associated with higher prob of higher bleaching (in north)
exp(0.99968)
## Central is associated with lower prob of higher bleaching (in habitat c)
exp(-2.29)
## South is associated with a lower prob of higher bleaching
exp(-7.51)
# Significant interactions due to F:Central (Lower) and F:Southern (lower)

exp(-9.1435)
exp(-0.7224)
exp(-0.3714)

emmeans(hughes.clmm, ~oSCORE|HABITAT+SECTOR, mode='prob')
emmeans(hughes.clmm1, ~oSCORE|HABITAT+SECTOR, mode='prob')
emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class')
emmeans(hughes.clmm1, ~HABITAT|SECTOR, mode='mean.class')
emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class') %>% pairs()
emmeans(hughes.clmm1, ~HABITAT|SECTOR, mode='mean.class') %>% pairs()

## newdata = emmeans(hughes.clmm, ~ HABITAT|SECTOR, mode='mean.class') %>% as.data.frame %>%
newdata = emmeans(hughes.clmm1, ~ HABITAT|SECTOR, mode='mean.class') %>% as.data.frame %>%
    mutate(across(c(mean.class, asymp.LCL, asymp.UCL), function(x) x-1)) 
newdata
ScoreBoundaries = data.frame(Score=factor(0:4), ymin=c(0:4), ymax=c(1:5))
ggplot(newdata) +
    geom_blank(aes(y=mean.class, x=HABITAT)) + 
    geom_hline(yintercept=1, linetype='dashed', size=0.1) +
    geom_hline(yintercept=2, linetype='dashed', size=0.1) +
    geom_hline(yintercept=3, linetype='dashed', size=0.1) +
    #geom_rect(data=ScoreBoundaries, aes(ymin=ymin, ymax=ymax, xmin=-Inf, xmax=Inf, fill=Score), alpha=0.2) +
    geom_pointrange(aes(y=mean.class, x=HABITAT, ymin=asymp.LCL, ymax=asymp.UCL)) +
    facet_grid(~SECTOR) +
    scale_y_continuous('Bleaching score', breaks=(0:4), labels=0:4, limits=c(0,4),expand=c(0,0)) +
    theme_bw() +
    theme(panel.spacing.y=unit(10,'pt'))


## Pairwise for habitat
## newdata = emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class') %>%
newdata = emmeans(hughes.clmm1, ~HABITAT|SECTOR, mode='mean.class') %>%
    pairs() %>% confint() %>% as.data.frame()

ggplot(newdata) +
    geom_hline(yintercept=0) + 
    geom_pointrange(aes(y=estimate, x=contrast, ymin=asymp.LCL, ymax=asymp.UCL, color=SECTOR),
                    position=position_dodge(width=0.5)) +
    facet_grid(~SECTOR) +
    coord_flip() +
    scale_y_continuous('Effect size')+
    theme_bw()

## Pairwise for year

## newdata = emmeans(hughes.clmm, pairwise~ fYear|Habitat+SectorThree, mode='mean.class')$contrasts %>%
##                                                                                   confint %>% as.data.frame
## ggplot(newdata) +
##     geom_hline(yintercept=0) + 
##     geom_pointrange(aes(y=estimate, x=Habitat, ymin=asymp.LCL, ymax=asymp.UCL, color=SectorThree),
##                     position=position_dodge(width=0.5)) +
##     facet_grid(~SectorThree) +
##     coord_flip() +
##     scale_y_continuous('Effect size')+
##     theme_bw()

```
Conclusions:

- bleaching severity is substantially higher in habitat U than either L, F or C, lower in L than F or C, and lower in F than C in the North during 2016.
- the patterns in 2016 Central and Southern are similar to those described above.
- these patterns hold again for 2017 Southern habitats and to a lesser degree for the Central habitats
- the 2017 Northern patterns are a little different (mainly driven by higher relativel bleaching severity of Habitat F and relatively lower severity of U).
- bleaching severity was substantially higher in 2016 than 2017 in the North for all habitats (yet for F to a lesser degree)
- bleaching severity was substantially lower in 2016 than 2017 in the Central for all habitats
- bleaching severity was subtly higher in 2016 than 2017 in the South for all habitats


```{r name, results='markdown', eval=FALSE}
hughes %>% dplyr::filter(fYear=='2016', SectorThree=='North') %>%
    group_by(Habitat) %>%
    count(oScore) %>%
    group_by(Habitat) %>%
    mutate(Prop = n/sum(n))
    ggplot() + geom_point(aes(oScore

hughes.clmm=ordinal::clmm(oScore ~ Habitat+(1|ReefID),
                          data=hughes %>% dplyr::filter(fYear=='2016', SectorThree=='North'))
summary(hughes.clmm)
emmeans(hughes.clmm, ~oScore|Habitat, mode='prob')
emmeans(hughes.clmm, ~Habitat, mode='mean.class')
emmeans(hughes.clmm, pairwise~Habitat, mode='mean.class')

```



</div>
