---
title: "GLMM example 9"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../public/resources/ws_style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../public/resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(broom.mixed) #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(ggeffects) #for effects plots in ggplotjk
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(DHARMa)    #for assessing dispersion etc
library(glmmTMB)    #for glmmTMB
library(performance) #for diagnostic plots
library(see)         #for diagnostic plots
library(ordinal)    #for ordinal models
```

<div class="HIDDEN">

```{r readDataP, results='markdown', eval=TRUE}
hughes = read_csv('../public/data/hughes_full.csv', trim_ws=TRUE)
glimpse(hughes)
```

| Bleaching Score | 2016 Equates to  | 2017 Equates to  |
|-----------------+------------------+------------------|
|               0 | No bleaching     | No bleaching     |
|               1 | 0-10% bleaching  | 0-10% bleaching  |
|               2 | 10-30% bleaching | 10-30% bleaching |
|               3 | 30-60% bleaching | 30-60% bleaching |
|               4 | > 60% bleaching  | 60-80% bleaching |
|               5 | -                | > 80% bleaching  |

# Data processing

- Make a categorical version of Year
- Combine Scores 4 and 5 into a single Score of 4 to be consistent across years
- Make Score an ordered categorical variable
- Generate a numeric version of categorical Score (as required for the `ocat` family.  Note this must be a 1 indexed numeric.
- Relevel Sector such that it is in the order North, Central, South
- Ensure that there is a categorical version of Reef
- Generate a numeric version of the categorical Reef 


```{r processDataP, results='markdown', eval=TRUE}
hughes = hughes %>%
    mutate(fYear=factor(Year),
           Score=ifelse(Score==5,4,Score),
           oScore = factor(Score, ordered=TRUE),
           nScore = as.numeric(factor(Score, ordered=TRUE)),
           SectorThree=factor(SectorThree, levels=c('North','Central','South')),
           fReef=factor(ReefID),
           nReef=as.numeric(fReef))
# now make a version that is just 2016
hughes <- hughes %>% filter(fYear==2016) %>%
    dplyr::select(REEF=ReefID, HABITAT=Habitat, SECTOR=SectorThree, SCORE=Score)
write_csv(hughes, file='../public/data/hughes.csv')
## hughes.colors = c('#FFFFFF', rev(heat.colors(length(levels(hughes$oScore))))[-1])
```
</div>

# Scenario

# Read in the data

```{r readData, results='markdown', eval=TRUE}
hughes = read_csv('../public/data/hughes.csv', trim_ws=TRUE)
glimpse(hughes)
```

| REEF   | HABITAT | SECTOR | SCORE |
|--------|---------|--------|-------|
| 09-357 | C       | North  | 4     |
| 09-357 | F       | North  | 4     |
| 09-357 | U       | North  | 3     |
|        |         |        |       |

# Data preparation

<div class="HIDDEN">

```{r processData, results='markdown', eval=TRUE}
hughes = hughes %>%
    mutate(oSCORE = factor(SCORE, ordered=TRUE),
           HABITAT = factor(HABITAT),
           SECTOR=factor(SECTOR, levels=c('North','Central','South')),
           REEF=factor(REEF))
hughes.colors = c('#FFFFFF', rev(heat.colors(length(levels(hughes$oSCORE))))[-1])
```
</div>

# Exploratory Data Analysis

<div class="HIDDEN">

```{r EDA2, results='markdown', eval=TRUE, fig.width=10, fig.height=10}
hughes.sum <- hughes %>%
    count(SECTOR,HABITAT,oSCORE) %>%
    group_by(SECTOR, HABITAT) %>%
    mutate(prop=prop.table(n),
           oSCORE=factor(oSCORE, levels=rev(levels(oSCORE))))

hughes.sum %>% head

ggplot(data=hughes.sum, aes(y=prop, x=HABITAT)) +
    geom_bar(stat='Identity', aes(fill=oSCORE), color='black') +
    facet_grid(~SECTOR) +
    ## scale_fill_manual('Bleaching score', values=rev(hughes.colors) ) +
    scale_fill_manual('Bleaching score', values=c(heat.colors(5)[-5], '#FFFFFF') ) +
    scale_y_continuous('Proportion of Reef', expand=c(0,0))+
    theme_bw() + 
    theme(panel.spacing.y=unit(10,'pt'))
```

</div>

# Fit models

<div class="HIDDEN">

```{r mod1Fit, results='markdown', eval=FALSE, cache=TRUE, results='hide'}

hughes.clmm=ordinal::clmm(oSCORE ~ HABITAT*SECTOR+(1|REEF), data=hughes)
allEffects(hughes.clmm) %>% plot
plot_model(hughes.clmm, type='eff', terms=c('HABITAT','SECTOR'))
plot_model(hughes.clmm, type='eff', terms=c('SECTOR','HABITAT'))
ggpredict(hughes.clmm) %>% plot

## predict(hughes.clmm)
## model=hughes.clmm
## coefs <- c(model$beta, unlist(model$ST))

## simulateResiduals(hughes.clmm, plot=TRUE)
## hughes.clmm=ordinal::clmm(oScore ~ Habitat*fYear*SectorThree+(1|ReefID), data=hughes)
summary(hughes.clmm)

exp(-9.1435)
exp(-0.7224)
exp(-0.3714)

emmeans(hughes.clmm, ~oSCORE|HABITAT+SECTOR, mode='prob')
emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class')
emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class')
emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class') %>% pairs()
emmeans(hughes.clmm, pairwise~Habitat, mode='mean.class')

newdata = emmeans(hughes.clmm, ~ HABITAT|SECTOR, mode='mean.class') %>% as.data.frame %>%
    mutate(across(c(mean.class, asymp.LCL, asymp.UCL), function(x) x-1)) 
newdata
ScoreBoundaries = data.frame(Score=factor(0:4), ymin=c(0:4), ymax=c(1:5))
ggplot(newdata) +
    geom_blank(aes(y=mean.class, x=HABITAT)) + 
    geom_hline(yintercept=1, linetype='dashed', size=0.1) +
    geom_hline(yintercept=2, linetype='dashed', size=0.1) +
    geom_hline(yintercept=3, linetype='dashed', size=0.1) +
    #geom_rect(data=ScoreBoundaries, aes(ymin=ymin, ymax=ymax, xmin=-Inf, xmax=Inf, fill=Score), alpha=0.2) +
    geom_pointrange(aes(y=mean.class, x=HABITAT, ymin=asymp.LCL, ymax=asymp.UCL)) +
    facet_grid(~SECTOR) +
    scale_y_continuous('Bleaching score', breaks=(0:4), labels=0:4, limits=c(0,4),expand=c(0,0)) +
    theme_bw() +
    theme(panel.spacing.y=unit(10,'pt'))


## Pairwise for habitat
newdata = emmeans(hughes.clmm, ~HABITAT|SECTOR, mode='mean.class') %>%
    pairs() %>% confint() %>% as.data.frame()

ggplot(newdata) +
    geom_hline(yintercept=0) + 
    geom_pointrange(aes(y=estimate, x=contrast, ymin=asymp.LCL, ymax=asymp.UCL, color=SECTOR),
                    position=position_dodge(width=0.5)) +
    facet_grid(~SECTOR) +
    coord_flip() +
    scale_y_continuous('Effect size')+
    theme_bw()

## Pairwise for year

newdata = emmeans(hughes.clmm, pairwise~ fYear|Habitat+SectorThree, mode='mean.class')$contrasts %>%
                                                                                  confint %>% as.data.frame
ggplot(newdata) +
    geom_hline(yintercept=0) + 
    geom_pointrange(aes(y=estimate, x=Habitat, ymin=asymp.LCL, ymax=asymp.UCL, color=SectorThree),
                    position=position_dodge(width=0.5)) +
    facet_grid(~SectorThree) +
    coord_flip() +
    scale_y_continuous('Effect size')+
    theme_bw()

```
Conclusions:

- bleaching severity is substantially higher in habitat U than either L, F or C, lower in L than F or C, and lower in F than C in the North during 2016.
- the patterns in 2016 Central and Southern are similar to those described above.
- these patterns hold again for 2017 Southern habitats and to a lesser degree for the Central habitats
- the 2017 Northern patterns are a little different (mainly driven by higher relativel bleaching severity of Habitat F and relatively lower severity of U).
- bleaching severity was substantially higher in 2016 than 2017 in the North for all habitats (yet for F to a lesser degree)
- bleaching severity was substantially lower in 2016 than 2017 in the Central for all habitats
- bleaching severity was subtly higher in 2016 than 2017 in the South for all habitats


```{r name, results='markdown', eval=FALSE}
hughes %>% dplyr::filter(fYear=='2016', SectorThree=='North') %>%
    group_by(Habitat) %>%
    count(oScore) %>%
    group_by(Habitat) %>%
    mutate(Prop = n/sum(n))
    ggplot() + geom_point(aes(oScore

hughes.clmm=ordinal::clmm(oScore ~ Habitat+(1|ReefID),
                          data=hughes %>% dplyr::filter(fYear=='2016', SectorThree=='North'))
summary(hughes.clmm)
emmeans(hughes.clmm, ~oScore|Habitat, mode='prob')
emmeans(hughes.clmm, ~Habitat, mode='mean.class')
emmeans(hughes.clmm, pairwise~Habitat, mode='mean.class')

```

</div>
